### 2015-09-01 Manuscript/SW_early_life_telomeres_SOM_v9.Rmd
Added longitudinal data. General updates.

**Table S1** Model selection table for analysis of factors affecting early life telomere length in juvenile Seychelles warblers. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. i = intercept.

@@

#Run the model
lm1 <- lmer(TLKB ~ Agemonths + TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

@@

#Model averaging
lm2 <- dredge(lm1)
t1 <- as.matrix(subset(lm2, delta<=6))
colnames(t1)[2:7] <- c('Sex', 'Body condition','Group size','Helper','Food av.','TQ')
getterms <- function(x)
  {
  temp <- c('i',names(which(!(is.na(x[2:7])))))
  return(paste(temp,collapse = " + "))
  }

tableS1 <- data.frame(Model = apply(t1,1,getterms),t1[,c('df','AICc','delta','weight')])
colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS1) <- NULL
tableS1$Model[tableS1$Model == ''] <- 'Null model'
kable(tableS1,digits=3)

@@

ddFigS2 <- ddply(temp,
                 .(LayYear),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

ddFigS2 <- ddFigS2[complete.cases(ddFigS2),]
ddFigS2$LayYear <- as.numeric(paste(ddFigS2$LayYear))

FigS2 <- ggplot(ddFigS2,
       aes(x = LayYear,
           y = TLKBmean)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  ylab('Telomere length (kb)') + 
  xlab('Lay Year') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

@@

**Figure S2** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers

>>>>

**Table S1** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler chicks. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. i = intercept.

@@

#Run the model
lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + BodyMass + Tarsus + Helper + (1|LayYear),
            data = chickall,REML=FALSE)

@@

#Model averaging
lm2 <- dredge(lm1)
t1 <- as.matrix(subset(lm2, delta<=6))
colnames(t1)[2:8] <- c('Sex', 'Body Mass','Helpers','Food av.','Non helpers','Tarsus length','TQ')

tableS1 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')])
colnames(tableS1) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tableS1) <- NULL
tableS1$Model[tableS1$Model == ''] <- 'Null model'
kable(tableS1,digits=3)

**Table S2** Model selection table for analysis of factors affecting early life telomere length in Seychelles warbler fledglings and subadults. Only models that remained in the top model set for model averaging (AICc $\leq$ 6 compared to the top model) are included here. i = intercept.

```{r Table s2}

#Run the model
lm1 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + BodyMass + Tarsus + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

#Model averaging
lm2 <- dredge(lm1)
t1 <- as.matrix(subset(lm2, delta<=6))
colnames(t1)[2:8] <- c('Sex', 'Body Mass','Helpers','Food av.','Non helpers','Tarsus length','TQ')

tables2 <- data.frame(Model = apply(t1,1,getterms,c(2:8)),t1[,c('df','AICc','delta','weight')])
colnames(tables2) <- c('Model','df','AICc', 'Delta AICc','Weight')
rownames(tables2) <- NULL
tables2$Model[tables2$Model == ''] <- 'Null model'
kable(tables2,digits=3)

@@

temp <- Loss
levels(temp$Ageclass) <- c('Chicks', 'Fledglings','Subadults')

ddFigS2 <- ddply(temp,
                 .(Ageclass),
                 summarise,
                 TROCmean = mean(TROC),
                 TROCse = se(TROC))

FigS2 <- ggplot(ddFigS2,
       aes(x = Ageclass,
           y = TROCmean)) +
  geom_pointrange(aes(min=TROCmean-TROCse,ymax=TROCmean+TROCse)) +
  ylab('Corrected telomere loss') + 
  xlab('') + 
  theme_classic() +
  ylim(0,0.002) +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

@@

**Figure S2** Telomere loss (mean $\pm$ s.e.) in juvenile Seychelles warblers. Telomere loss a standardised and corrected value (see methods for details) based on rates of lost between the sampling event of birds as a juvnile, and the first available sampling as an adult. Values above zero indicate telomere shortening in early life.

```{r Figure S3,fig.width=4,fig.height=4}

ddFigS3 <- ddply(temp,
                 .(LayYear),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB))

ddFigS3 <- ddFigS3[complete.cases(ddFigS3),]
ddFigS3$LayYear <- as.numeric(paste(ddFigS3$LayYear))

FigS3 <- ggplot(ddFigS3,
       aes(x = LayYear,
           y = TLKBmean)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  ylab('Telomere length (kb)') + 
  xlab('Lay Year') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16))

FigS3

**Figure S3** Cross-sectional juvenile telomere length (mean $\pm$ s.e.) across birth years in Seychelles warblers

### 2015-09-01 Manuscript/SW_early_life_telomeres_tables_and_figures v9.Rmd
Added longitudinal data. General updates.

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.width= 5,
               fig.height= 5,
               fig.show='asis',
               dev='png')

@@

**Figure 1** Factors affecting telomere length in juvenile Seychelles warblers. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable. Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av = annual food availability, TQ = territory quality). **B** Annual food availability and **C** body condition in relation to telomere length. Points and error bars in **A** are mean $\pm$ 95% confidence interval telomere lengths for each year or group size (means are used for plotting purposes only), and the lines represent fitted values and from linear regressions.

**Figure 2** Juvenile telomere length and lifespan in Seychelles warblers. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a generalised linear model with a poisson error structure and lifespan as the response variable. Only juveniles that survived their first year of life were included in the model. Numbers in brackets are the relative importance of each term in the top model set (see main text for details). Food av = annual food availability, TL = telomere length TQ = territory quality. **B** Raw telomere length and annual food availability in relation to lifespan. Annual food availability was modeled as a covariate (see main text for details) but is plotted as a factor here for visualisation purposes. The lines represent fitted values from a Poisson regression.

@@

```{r Figure 1,fig.width=8,fig.height=9}

###########
#Fig 1A

#Run the model
lm1 <- lmer(TLKB~ TQspace + Insect + GroupSize + Sex + Condition + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

lm1 <- standardize(lm1)

#Model averaging
lm2 <- dredge(lm1)
lmavg <- model.avg(lm2,subset = delta <=6)
myimp <- importance(lmavg) #Relative importance of each explanatory variable

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.Condition'] <- 'Body condition'
nams[nams == 'z.GroupSize'] <- 'Local density'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

Fig1A <- forestplot(table1) +
  annotate('text',x=0.5,y=4,label='A')

ddFig1B <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig1B <- ddFig1B[complete.cases(ddFig1B),]
ddFig1B$Insect <- as.numeric(paste(ddFig1B$Insect))

Fig1B <- ggplot(ddFig1B,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Annual food availability') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=F) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=8,y=12,label='B')

Fig1C <- ggplot(FlSAall,
                aes(x = Condition,
                    y = TLKB)) +
    geom_point() +
  ylab('Telomere length (kb)') + 
  xlab('Body Condition') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(method=lm,col='black',se=F) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=-2.5,y=15,label='C')

multiplot(Fig1A,Fig1B,Fig1C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))

@@

```{r Figure 2}

x <- summary(lm(TLKB~Insect,data= FlSAall))$resid
xf <- ifelse(x>median(x),'Long','Short')

s1 <- survfit(Surv(time = Lifespan+1, event = Died)~xf, data = FlSAall)
Fig2 <- ggsurv(s1,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion Survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  scale_colour_manual(name = 'Residual TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  theme(legend.title = element_text('Telomere length'))

Fig2

@@

>>>>

#Set some options
options(na.action='na.fail')
opts_knit$set(root.dir = getwd())
opts_chunk$set(echo=FALSE,
               warning=FALSE,
               message=FALSE,
               fig.path='../Figures/',
               fig.show='asis',
               dev='png')

@@

**Figure 3** Telomere length and age in juvenile Seychelles warblers. Points and error bars represent mean and 95% confidence intervals, respectively.

**Figure 2** Factors affecting telomere length in Seychelles warbler chicks. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable. Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av. = annual food availability, TQ = territory quality). **B** Presence/absence of helpers and **C** body mass in relation to raw telomere length. Points and error bars in **B** are mean $\pm$ 95% confidence interval telomere lengths for each group, and the line and shaded areas from **C** represent fitted values and 95% confidence limits from a linear regression.

**Figure 3** Factors affecting telomere length in juvenile Seychelles warblers. **A** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a linear model with juvenile telomere length as the response variable. Numbers in brackets are the relative importance of each term in the top model set (see main text for details; Food av. = annual food availability, TQ = territory quality). **B** Raw telomere length in relation to annual food availability Points and error bars in **B** are mean $\pm$ 95% confidence interval telomere lengths for each year (means are used for plotting purposes only), and the line and shaded areas represent fitted values and 95% confidence limits from a linear regression.

**Figure 4** Juvenile telomere length and survival in chick (**A,B**) and juvenile (**C,D**) Seychelles warblers. **A,C** Model averaged estimates and 95% confidence intervals (points and error bars, respectively) for all explanatory terms used in a parametric survival model (see main text for details). Food av = annual food availability, Tel. length = telomere length. **B,D** Kaplan-Meier curves showing the relationship between telomere length and survival. Telomere length was modelled as a covariate, but is binned into groups here for visualisation purposes (long and short = greater than or less than median telomere length, repectively).

@@

```{r Figure 1,fig.width=5,fig.height=4}

temp <- dd

ddFig1 <- ddply(temp,
                 .(AgemonthF),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = se(TLKB)*1.96)

Fig1 <- ggplot(ddFig1,
       aes(x = AgemonthF,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Age (months)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse))

Fig1

@@

```{r Figure 2,fig.width=8,fig.height=9}

###########
#Fig 2A

chickall <- subset(chickall,BodyMass>5)
#Run the model
lm2 <- lmer(TLKB ~ TQspace + Insect + BodyMass+ Tarsus + NonHelper + Sex + Helper + (1|LayYear),
            data = chickall,REML=FALSE)

lm2 <- standardize(lm2)

#Model averaging
lm3 <- dredge(lm2)
lmavg <- model.avg(lm3,subset = delta <=6)
myimp <- importance(lmavg) #Relative importance of each explanatory variable

#Put it all into a nice table
table2 <- imptable(lmavg,myimp)
table2 <- signif(table2,2)

#Sort out row/col names
nams <- rownames(table2)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

rownames(table2) <- nams
colnames(table2) <- c('Estimate','SE','2.5%','97.5%','RI')

table2 <- data.frame(table2,myx=row.names(table2))

Fig2A <- forestplot(table2) +
  annotate('text',x=0.75,y=4,label='A')

######################################
temp <- chickall
temp$Helper <- ifelse(temp$Helper>0,'Present','Absent')

ddFig2B <- ddply(temp,
                 .(Helper),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))

Fig2B <- ggplot(ddFig2B,
       aes(x = Helper,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Helpers') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=0.65,y=9,label='B')

Fig2C <- ggplot(chickall,
       aes(x = BodyMass,
           y = TLKB)) +
  ylab('Telomere length (kb)') + 
  xlab('Body mass (g)') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  geom_point() +
  stat_smooth(method = 'lm',col = 'black') +
  annotate('text',x=10.5,y=15,label='C')

multiplot(Fig2A,Fig2B,Fig2C,layout=matrix(c(1,1,1,1,1,1,2,3,2,3), nrow=5, byrow=TRUE))

@@

**Figure 3**

```{r Figure 3,fig.width=5,fig.height=9}
###########
#Fig 3A

#Run the model
lm3 <- lmer(TLKB ~ TQspace + Insect + NonHelper + Sex + Tarsus + BodyMass + Helper + (1|LayYear),
            data = FlSAall,REML=FALSE)

lm3 <- standardize(lm3)

#Model averaging
lm4 <- dredge(lm3)
lmavg <- model.avg(lm4,subset = delta <=6)
myimp <- importance(lmavg) #Relative importance of each explanatory variable

#Sort out row/col names
nams <- rownames(table1)
nams[nams == 'c.Ageclass'] <- 'Age class'
nams[nams == 'z.BodyMass'] <- 'Body mass'
nams[nams == 'z.Tarsus'] <- 'Tarsus'
nams[nams == 'z.NonHelper'] <- 'Non helpers'
nams[nams == 'c.Sex'] <- 'Sex'
nams[nams == 'z.Insect'] <- 'Food av.'
nams[nams == 'z.TQspace'] <- 'TQ'
nams[nams == 'z.Helper'] <- 'Helpers'

Fig3A <- forestplot(table1) +
  annotate('text',x=0.75,y=2,label='A')

ddFig3B <- ddply(temp,
                 .(Insect),
                 summarize,
                 TLKBmean = mean(TLKB),
                 TLKBse = 1.96*se(TLKB))
ddFig3B <- droplevels(ddFig3B[complete.cases(ddFig3B),])
ddFig3B$Insect <- as.numeric(paste(ddFig3B$Insect))

Fig3B <- ggplot(ddFig3B,
       aes(x = Insect,
           y = TLKBmean)) +
  ylab('Telomere length (kb)') + 
  xlab('Annual food availability') +
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16)) +
  stat_smooth(data=FlSAall,aes(x=Insect,y=TLKB),method=lm,col='black',se=T) +
  geom_pointrange(aes(ymin=TLKBmean-TLKBse,ymax=TLKBmean+TLKBse)) +
  scale_colour_manual(values = wes_palette("Royal1")) +
  annotate('text',x=2,y=10,label='B')

multiplot(Fig3A,Fig3B)

**Figure 4**

```{r Figure 4,fig.height=8,fig.width=8}
s1 <- with(subset(dd,Ageclass == 'FL'),flexsurvreg(Surv(Lifespan+1,Died)~TLKB, dist="lnorm"))

s1 <- with(chicks,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+BodyMass+TLKB+Helper, dist="lnorm"))
t1 <- data.frame(s1$res)
t1 <- t1[3:5,]
t1 <- t1[order(rownames(t1)),]
rownames(t1) <- c('Body mass','Helper','Tel. length')

Fig4A <- ggplot(t1,aes(x = rownames(t1),y = est)) +
  geom_point() + 
  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
    theme_classic() +
  ylab('Estimate') + 
  xlab('') + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90,size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  geom_abline(intercept = 0,slope = 0,lty=2) +
  annotate('text',x=0.6,y=0.4,label='A')

s2 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = chicks)
Fig4B <- ggsurv(s2,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.text = element_text(size = 12),
        legend.title = element_blank()) +
  scale_colour_manual(name = 'TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  annotate('text',x=0.5,y=1.1,label='B')

###########################################

s2 <- with(FlSAall,flexsurvreg(Surv(Lifespan+1,Died)~TLKB+Insect, dist="lnorm"))
t4 <- data.frame(s2$res)
t4 <- t4[3:4,]
t4 <- t4[order(rownames(t4)),]
rownames(t4) <- c('Food av.','Tel. length')

Fig4C <- ggplot(t4,aes(x = rownames(t4),y = est)) +
  geom_point() + 
  geom_pointrange(aes(ymin = L95.,ymax = U95.)) +
    theme_classic() +
  ylab('Estimate') + 
  xlab('') + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90,size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12)) +
  geom_abline(intercept = 0,slope = 0,lty=2) +
  annotate('text',x=0.6,y=0.15,label='C')

s4 <- survfit(Surv(time = Lifespan+1, event = Died)~TLF, data = FlSA)
Fig4D <- ggsurv(s4,plot.cens=F) + 
  theme_classic() +
  ylab('Proportion survived') + 
  xlab('Age (years)') + 
  theme_classic() +
  theme(axis.text.x = element_text(size = 14), 
        axis.text.y = element_text(size = 14), 
        axis.title.y = element_text(size = 16,vjust=0.8),
        axis.title.x = element_text(size = 16),
        legend.position = c(0.8,0.8),
        legend.text = element_text(size = 12),
        legend.title = element_blank()) +
  scale_colour_manual(name = 'TL',
                      values = c('darkred','red')) +
  guides(linetype=F) +
  annotate('text',x=0.5,y=1.1,label='D')

multiplot(Fig4A,Fig4B,Fig4C,Fig4D,layout=matrix(c(1:4), nrow=2, byrow=FALSE))

### 2015-09-01 Manuscript/SW_early_life_telomeres_v9.Rmd
Added longitudinal data. General updates.

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: MyBib.bib
csl: ProceedingsB.csl
---
```{r load and clean, include = F}

@@

Lewis G. Spurgin^1^,  Kathryn L. Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^2^, Terry Burke^3^, Hannah, L. Dugdale^3^,  and David S. Richardson^1,4^.

1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
2. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
3. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
4. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

Telomere lengths were measured from `r nrow(FlSA)` samples taken cross-sectionally from birds aged between two and twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults). We chose to focus on juveniles and subadults rather than nestlings, because i) we have juvenile samples spanning a range of years, whereas nestling samples are predominantly from two early years, and ii) telomere length during the nestling phase will largely be a product of inheritance and growth, whereas in juveniles and subadults we expect telomere length to incorporate the impact of early-life experiences. We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We first tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, age class (juvenile or subadult), territory quality, yearly food availability and local density (measured as the number of individuals in the juvenile's natal social group [@Brouwer2012]). Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

@@

@@

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). We found no significant difference in telomere length between age classes (juvenile vs subadult), or between sexes (Figure 1A).

Early-life telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 1B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 1A; Table S1). The best supported model with early-life telomere length as the response variable contained annual food availability and body condition as explanatory variables (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect+GroupSize+Condition,data=FlSAall))$r.squared)`). Early-life telomere length  was positively related to body condition (Figure 1C).

@@

FlSAall$LayYear <- factor(FlSAall$LayYear)
summary(glmer(Lifespan~TLKB+Insect+Condition+(1|LayYear),data=FlSAall,family=poisson))

>>>>

---
output:
  word_document:
    fig_caption: yes
    highlight: null
bibliography: MyBib.bib
csl: JAE.csl
---
```{r load and clean, include = F}

@@

Lewis G. Spurgin^1,2^,  Kat Bebbington^1^, Eleanor A. Fairfield^1^, Jan Komdeur^3^, Terry Burke^4^, Hannah, L. Dugdale^2^,  and David S. Richardson^1,5^.

1. School of Biological Sciences, University of East Anglia, Norwich Research Park, NR4 7TJ, United Kingdom
2. Department of Zoology, Edward Grey Institute, University of Oxford, Oxford, UK
3. Behavioural Ecology and Self-organization Group, Centre for Ecological and Evolutionary Studies, University of Groningen, Groningen, The Netherlands
4. Department of Animal and Plant Sciences, NERC Biomolecular Analysis Facility, University of Sheffield, Sheffield, UK
5. Nature Seychelles, Roche Caiman, Mahé, Republic of Seychelles

@@

Telomere lengths were measured from `r nrow(FlSA)+nrow(chicks)` samples taken cross-sectionally from birds aged less than twelve months, between `r min(as.numeric(paste(juv$LayYear)))` and `r max(as.numeric(paste(juv$LayYear)))`. Of these birds, `r nrow(chicks)` were nestlings less than two months old, `r nrow(subset(FlSA,Ageclass=='FL'))` were aged between two and six months (juveniles), and `r nrow(subset(FlSA,Ageclass=='SA'))` were aged seven to twelve months (subadults). We measured absolute telomere quantity per diploid genome using a quantitative PCR (qPCR) assay  (full details in [@Barrett2012]).

#####Statistical analyses
We performed all analyses using R version 3.0.1 [@RDevelopmentCoreTeam2011]. We tested whether telomere length varied within the first year of life using one-way ANOVA, with telomere length as the response variable and age class (chick, juvenile or subadult) as a fixed factor. As telomere length varied substantially among age classes (see results) we carried out all subsequent analyses seperately for each age class.

We tested tested whether early-life environmental variation influences telomere length using linear mixed models, with telomere length as the response variable. As explanatory variables we included body condition (calculated as deviation from a regression line of mass on tarsus length; [@VandeCrommenacker2011a]), sex, territory quality, yearly food availability, and the number of helping and non-helping subordinates present in the natal territory  ([@Brouwer2012]). Lay Year was included as a random factor, in order to account for any annual variation in telomere length not explained by yearly food availability. We created a full model containing all of the above terms, and a top model set was then defined, containing all models with AICc $\leq$ 6 compared to the best supported model [@Burnham2011]. We report model averaged coefficients, confidence intervals, and 'relative importance', which reflects the relative weights of each predictor variable across the top model set. Model averaging was carried out using the MuMIn package (version 1.10.5) in R [@Barton2012].

@@

For a subset of juveniles (n = `r nrow(Loss)` individuals) we had longitudinal data (i.e samples from individuals caught both as a juvenile an adult). For these individuals we calculated a rate of telomere loss between the juvenile sample and earliest adult sample by subtracting the adult telomere length from the juvenile telomere length and dividing this difference by the length of time (in days) between sampling events. To account for regression to the mean effects, we applied a correction based on correlations among samples within individuals following Verhulst *et al.* [-@Verhulst2013].

@@

Mean ($\pm$ standard error) early-life telomere length was `r round2(mean(FlSAall$TLKB))` $\pm$ `r round2(se(FlSAall$TLKB))` kb. This is only marginally higher than mean telomere length for all adult birds (`r round2(mean(subset(dd,Age>0)$TLKB))` $\pm$ `r round2(se(subset(dd,Age>0)$TLKB))` kb), and lower than the mean for one year-old birds based on cross-sectional data (`r round2(mean(subset(dd,Age==1)$TLKB))` $\pm$ `r round2(se(subset(dd,Age==1)$TLKB))` kb; Figure S1). However, when we considered ageclasses seperately, we found substantial age-related variation in telomere length within the first year of life (F = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,4])`; *P* = `r round2(anova(lm(TLKB~Ageclass,data = juv))[1,5])`; Fig. 1). This cross-sectional data revealed that chicks have the longest telomeres, but that subsadults and adults had longer telomeres than fledglings (Fig. 1). However, a longitudinal analysis showed that chicks, fleglings nd subadults all lost telomeres early in life, with no differences in rates of telomere shortening among age classes (F = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,4])`; *P* = `r round2(anova(lm(TROC~Ageclass,data = Loss))[1,5])`; Fig. S2). This suggests that the apparent increase in telomere length between juveniles and subadults in the corss-secitonal data was a result of the sampling, rather than telomere elongation.

In chicks, telomere length did not vary significantly among birth years (Fig. 2A, Fig. S2A).

Juvenile telomere length differed substantially among birth years (Figure S2), with longer telomeres in years where insect abundance was high (Figure 2B). Model averaging confirmed this strong, positive effect of yearly food availability on telomere length (Figure 2A; Table S1). The best supported model with juvenile telomere length as the response variable contained annual food availability as the only explanatory variable (Table S1; R^2^ = `r round2(summary(lm(TLKB~Insect,data=FlSAall))$r.squared)`).

@@
